---
title: "Checkpoint"
output: html_document
date: "2024-07-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(factoextra)
library(patchwork)
library(cluster)
```


## Loading, Cleaning, and Scaling Lung Data
```{r lung data for elbows}
lung_meta40 <- read.csv("./lung_meta40.csv")

#removing all columns except the k-value ones 
new_lung_meta40 = subset(lung_meta40, select = -c(image_id, patient_id, gender, mhcII_status, age_at_diagnosis, stage_at_diagnosis, stage_numeric, pack_years, survival_days, survival_status, cause_of_death, adjuvant_therapy, time_to_recurrence_days, recurrence_or_lung_ca_death, totalCell))

# removing all the na entries 
new_lung_meta40 = na.omit(new_lung_meta40)

# scaling the data and further selecting columns 
new_lung_meta40 = as.data.frame(scale(new_lung_meta40)) |>
  select(-c(X, p_CK, p_CD8, p_CD14, p_Other, p_CD19, p_CD4)) 

```

## Assessing Number of Clusters
```{r lung number clusters}
# width sum of squares method (elbow plot)
wss_lung = fviz_nbclust(new_lung_meta40, kmeans, method = "wss")

# silhouette method
silhouette_lung = fviz_nbclust(new_lung_meta40, kmeans, method = "silhouette")

# gap statistic method 
gap_stat_lung = fviz_nbclust(new_lung_meta40, kmeans, method = "gap_stat")
wss_lung + silhouette_lung + gap_stat_lung

```
Based on these results, we are planning to use two clusters for K-means and hierarchical clustering (using Ward distance).

## Performing PCA
```{r pca-lung}
pca_lung40 <- prcomp(new_lung_meta40)

## view cumulative proportion of variance
summary(pca_lung40)
```
We will be using five principal components, which has a cumulative proportion of 0.627.

## Performing Hierarchical Clustering
```{r lung number}
set.seed(42)
# hierarchical clustering using Ward distance and first 5 principle components
hier_clust <- hcut(pca_lung40$x[,1:5], k = 2)
## dendrogram
plot(hier_clust)
## clusters drawn on dendrogram
rect.hclust(hier_clust, k = 2)
## silhouette scores for hierarchical clustering
fviz_silhouette(hier_clust)
```

## Initial EDA on Clusters Produced Using Hierarchical Method
```{r }
# Create data frame for EDA
lung_meta40_eda <- lung_meta40 |>
  filter(!is.na(p_CK) & !is.na(p_CD8) & !is.na(p_CD14) & !is.na(p_Other) & !is.na(p_CD19) & !is.na(p_CD4)
         & !is.na(k_CK) & !is.na(k_CD8) & !is.na(k_CD14) & !is.na(k_Other) & !is.na(k_Other) & !is.na(k_CD19)
         & !is.na(k_CD4) & !is.na(k_CK_CD8) & !is.na(k_CK_CD14) & !is.na(k_CK_Other) & !is.na(k_CK_CD19) & 
           !is.na(k_CK_CD4) & !is.na(k_CD8_CD14) & !is.na(k_CD8_Other) & !is.na(k_CD8_CD19) & !is.na(k_CD8_CD4)
         & !is.na(k_CD14_Other) & !is.na(k_CD14_CD19) & !is.na(k_CD14_CD4) & !is.na(k_Other_CD19) & 
           !is.na(k_Other_CD4) & !is.na(k_CD19_CD4))
lung_meta40_eda$cluster_id <- hier_clust$cluster

# count number of patients in each cluster using hierarchical
lung_meta40_eda |>
  group_by(cluster_id) |>
  count()

# create data frame containing medians for all numeric values in each cluster produced using hierarchical clustering
data <- lung_meta40_eda[, sapply(lung_meta40_eda, is.numeric)]
cluster_meds <- data |>
  group_by(cluster_id) |>
  summarize(across(everything(), median, na.rm = TRUE))
```
Our next step will be to produce a heatmap using these two clusters.

## Other Directions
We have also begun looking into spectral clustering for non-linear data. Using this method, we determined that we should use four clusters. Using code from [this tutorial](https://rpubs.com/gargeejagtap/SpectralClustering), we have produced some initial results shown below.
```{r}
# Spectral Clustering
# https://rpubs.com/gargeejagtap/SpectralClustering
## Create Similarity matrix of Euclidean distance between points
S <- as.matrix(dist(pca_lung40$x[,1:5]))


## Create Degree matrix
D <- matrix(0, nrow=nrow(pca_lung40$x[,1:5]), ncol = nrow(pca_lung40$x[,1:5])) # empty nxn matrix

for (i in 1:nrow(pca_lung40$x[,1:5])) {
  
  # Find top 10 nearest neighbors using Euclidean distance
  index <- order(S[i,])[2:11]
  
  # Assign value to neighbors
  D[i,][index] <- 1 
}

# find mutual neighbors
D = D + t(D) 
D[ D == 2 ] = 1

# find degrees of vertices
degrees = colSums(D) 
n = nrow(D)


## Compute Laplacian matrix
# Since k > 2 clusters (3), we normalize the Laplacian matrix:
laplacian = ( diag(n) - diag(degrees^(-1/2)) %*% D %*% diag(degrees^(-1/2)) )
# eigen(laplacian)

which.max(diff(sort(eigen(laplacian)$values, decreasing = F))) + 1


## Compute eigenvectors
eigenvectors = eigen(laplacian, symmetric = TRUE)
n = nrow(laplacian)
eigenvectors = eigenvectors$vectors[,(n - 2):(n - 1)]


set.seed(1748)
## Run Kmeans on eigenvectors
sc = kmeans(eigenvectors, 4)


## Pull clustering results
sc_results = cbind(pca_lung40$x[,1:5], cluster = as.factor(sc$cluster))
head(sc_results)

as.data.frame(sc_results) |>
  group_by(cluster) |>
  count()

ggplot(data = sc_results, aes(x = PC1, y = PC2)) + aes(color = factor(cluster)) + geom_point()

lung_meta40_eda_spectral <- lung_meta40 |>
  filter(!is.na(p_CK) & !is.na(p_CD8) & !is.na(p_CD14) & !is.na(p_Other) & !is.na(p_CD19) & !is.na(p_CD4)
         & !is.na(k_CK) & !is.na(k_CD8) & !is.na(k_CD14) & !is.na(k_Other) & !is.na(k_Other) & !is.na(k_CD19)
         & !is.na(k_CD4) & !is.na(k_CK_CD8) & !is.na(k_CK_CD14) & !is.na(k_CK_Other) & !is.na(k_CK_CD19) & 
           !is.na(k_CK_CD4) & !is.na(k_CD8_CD14) & !is.na(k_CD8_Other) & !is.na(k_CD8_CD19) & !is.na(k_CD8_CD4)
         & !is.na(k_CD14_Other) & !is.na(k_CD14_CD19) & !is.na(k_CD14_CD4) & !is.na(k_Other_CD19) & 
           !is.na(k_Other_CD4) & !is.na(k_CD19_CD4))

clusters <- sc_results[, 6]

lung_meta40_eda_spectral$cluster_id <- clusters
lung_meta40_eda_spectral |>
  group_by(cluster_id) |>
  count()

data <- lung_meta40_eda_spectral[, sapply(lung_meta40_eda_spectral, is.numeric)]
cluster_meds_spectral <- data |>
  group_by(cluster_id) |>
  summarize(across(everything(), median, na.rm = TRUE))

fviz_silhouette(silhouette(sc$cluster, dist(eigenvectors)))
```

